Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-07-04 paper/manuscript.tex

\newcommand{\versionnumber}{2.0} % you should update the minor version number in preprints and major version number of submissions.
\newcommand{\githubrepository}{\url{https://github.com/alchemistry/alchemical-best-practices}} %this should be the main github repository for this article
\newcommand{\expect}[1]{\left\langle{#1}\right\rangle}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]}
\author[1*]{Antonia S. J. S. Mey}
\author[2]{Bryce K. Allen}
\author[3]{Hannah E. Bruce Macdonald}
\author[3*]{John D. Chodera}
\author[9]{Vytautas Gapsys}
\author[9]{David F. Hahn}
\author[1,10]{Maximilian Kuhn}
\author[1]{Julien Michel}
\author[4*]{David L. Mobley}
\author[5]{Levi N. Naden}
\author[6]{Samarjeet Prasad}
\author[2,7]{Andrea Rizzi}
\author[1]{Jenke Scheen}
\author[11]{David R. Slochower}
\author[8*]{Michael R. Shirts}
\author[9]{Gary Tresadern}
\author[2]{Huafeng Xu}
%
\affil[1]{EaStCHEM School of Chemistry, David Brewster Road, Joseph Black Building, The King's Buildings, Edinburgh, EH9 3FJ, UK}
\affil[2]{Silicon Therapeutics, Boston, MA, USA}
\affil[3]{Computational and Systems Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York NY, USA}
\affil[4]{Departments of Pharmaceutical Sciences and Chemistry, University of California, Irvine, Irvine, USA}
\affil[5]{Molecular Sciences Software Institute, Blacksburg VA, USA}
\affil[6]{National Institutes of Health, Bethesda, MD, USA}
\affil[7]{Tri-Institutional Training Program in Computational Biology and Medicine, New York, NY, USA}
\affil[8]{University of Colorado Boulder, Boulder, CO, USA}
\affil[9]{Computational Chemistry, Janssen Research \& Development, Turnhoutseweg 30, Beerse B-2340, Belgium}
\affil[10]{Cresset, Cambridgeshire, UK}
\affil[11]{Vertex Pharmaceuticals, San Diego, CA, USA}

@@

\paragraph{Conserved binding site waters can play an important role in binding free energies}
Binding site water molecules may form water mediated protein-ligand interactions which can pose challenges whenever exchange with bulk water is slow compared to simulation timescales. This happens typically in buried binding sites. Overlaying multiple protein X-ray structures can identify conserved or additional water molecules that can be useful to include in calculations. In cases where water molecules are known to play an important role in the binding, software implementations that use water sampling facilitated by Grand Canonical Monte Carlo methods may be useful~\cite{michel2010prediction}. Other tools such as WaterMap or open source equivalents (SSTMap, GIST, and others) can be used to define water structure for systems with no experimental evidence of water sites~\cite{wang2011ligand}. Well-known protein systems with water mediated ligand interactions are for example: HSP90 which formed part of the D3R grand challenge 2015~\cite{mey2016blinded}, A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, ~\cite{deflorian2020accurate}, and others~\cite{michel2009energetics}.

@@

An absolute binding free energy calculation yields the standard free energy of formation of the molecular complex from the separated receptor and ligand species. Hence, as discussed in \ref{sec:non-covalent-molecular-binding}, it must depend on the standard state concentration $C^\circ$ (here $1 \, {\rm M} = 1 \, {\rm molecule}/1660$~\r{A}$^3$) and on the definition of the complex, that is on the specific binding pose for which the binding free energy is requested. Omitting the standard state concentration would yield an undefined result that would vary with the choice of concentration units. A similar vague estimate would result from omitting the binding pose definition that, for example, could equally describe binding to a specific site of the receptor or unspecific binding throughout the receptor surface depending on the extent of conformational sampling (see Sec.~\ref{sec:weak-binders}).

In the statistical mechanics formulation of Gilson et al.~\cite{gilson1997statisticalthermodynamic}, described in Section \ref{sec:non-covalent-molecular-binding}, the standard state concentration appears explicitly in Eq.~(\ref{eq:DG0-binding-constant-def}) and the binding pose is defined by the indicator function $I(\zeta)$ that characterizes the partition function of the complex; Eq.~(\ref{eq:intra-zRL-def}).  The role of the standard state and binding site definitions are particularly evident when expressing Eq.~(\ref{eq:DG0-binding-constant-def}) in terms of an ensemble average. To do so, we express the integrals in the numerator and denominator of Eq.~(\ref{eq:DG0-binding-constant-def}) over the same set of degrees of freedom. Note that the intramolecular configurational partition function of the complex, $z_{RL(v)}$, includes six more degrees of freedom--the position and orientation of the ligand relative to the receptor collectively described by the variable $\zeta$--than the product of the intramolecular partition functions, $z_{R(v)} z_{L(v)}$, of the separated receptor and ligand. To correct this discrepancy, we multiply and divide the right-hand side of Eq.~(\ref{eq:DG0-binding-constant-def})  by the quantity
\begin{equation} \label{eq:Vsite-def}
    V_{\rm site} \Omega = \int d\zeta I(\zeta) \, ,
\end{equation}
%\Omega is in my opinion a better symbol of the orientational extent than \xi_L. Lowercase greek letters are usually used
%for dynamical variables rather than integrated quantities
which measures the extent of the binding pose macrostate, where $V_{\rm site}$ corresponds to the spatial extent and $\Omega$ the orientational extent. The right-hand side of Eq.~(\ref{eq:Vsite-def}) combined with the integrals $z_{R(v)}$ and $z_{L(v)}$ yields a partition function similar to Eq.~(\ref{eq:intra-zRL-def}) but corresponding to a complex in which ligand and receptor are uncoupled~\cite{gallicchio2011recent} as if they were at infinite distance apart. The result is
\begin{equation}\label{eq:DG0-binding-constant-def-average}
K_b = C^\circ V_{\rm site} \frac{\Omega}{8 \pi^2} \langle e^{-\beta \Delta \Psi}  \rangle_0 
\end{equation}
where, similarly to Eq.~(\ref{eq:zratio-solv-average}),
\begin{equation}\label{eq:pert-energy-def}
\Delta \Psi(x_R, x_L, \zeta) =  \Psi_v(x_R, x_L, \zeta) - \left[  \Psi_v(x_R) + \Psi(x_L) \right]
\end{equation}
is the solvent-averaged binding energy of the complex when receptor and ligands are in intramolecular configurations $x_R$ and $x_L$, the ligand is in position and orientation $\zeta$ relative to the receptor (the uncoupled potential energy does not depend on  the position and orientation of the ligand relative to the receptor), and $\langle \ldots \rangle_0$ represents an ensemble average in the uncoupled ensemble. Finally, note that $\Omega = 8 \pi^2$ when the binding pose indicator function does not depend on the orientation of the ligand and Eq.~(\ref{eq:DG0-binding-constant-def-average}) simplifies to
\begin{equation}\label{eq:DG0-binding-constant-def-average-noOmega}
K_b = C^\circ V_{\rm site} \langle e^{-\beta \Delta \Psi}  \rangle_0 
\end{equation}
which, expressed in free energy units, yields
\begin{eqnarray}\label{eq:DG0-def-Vsite}
  \Delta G^\circ_b &=&  - k_B T \ln K_b = - k_B T \ln C^\circ V_{\rm site} - k_B T \ln \langle e^{-\beta \Delta \Psi}  \rangle_0 \nonumber \\
{\rm \ }  &=& \Delta G^\circ_{b,{\rm ideal}} + \Delta G_{b,{\rm exc.}}
\end{eqnarray}

The term $\Delta G^\circ_{b,{\rm ideal}} = - k_B T \ln C^\circ V_{\rm site}$ in Eq.~(\ref{eq:DG0-def-Vsite}) is interpreted as the \emph{ideal} component of the standard binding free energy, that is the binding free energy that would be measured in the absence of ligand-receptor interactions ($\Delta \Psi = 0$) and when the ligand's concentration is $C^\circ$. The ideal term is responsible for the dependence of the standard binding free energy on the standard state concentration. The arguably inappropriate term ``standard state correction'' is sometimes used in the literature to refer to $\Delta G^\circ_{b,{\rm ideal}}$. As seen above, the ideal term is a fundamental aspect of the statistical mechanics theory of molecular binding rather than a mere correction to an inexact formulation.

Conversely, the term $\Delta G_{b,{\rm exc.}} = - k_B T \ln \langle e^{-\beta \Delta \Psi}  \rangle_0 $ is the \emph{excess} component of the standard free energy of binding, which is the term that depends on intermolecular interactions and the specific chemical composition of the system. The excess component is the target of the alchemical molecular simulations discussed here, in contrast to the ideal term that is generally evaluated analytically~\cite{boresch2003absolute}.

\paragraph{Handling the standard state in free energy calculations.}
As discussed in Sections \ref{sec:theory-solvation} and \ref{sec:theory-partitioning}, the handling of the standard state in solvation and solvent partitioning free energy calculations is typically straightforward. Eq.~(\ref{eq:DG0-solvation-def}) identifies the ideal and excess components of the standard solvation free energy in terms of the standard gas pressure $P^\circ$, standard solution concentration $C^\circ$, and ensemble averages typically evaluated by alchemical free energy simulations. These relations provide simple recipes to relate computational estimates to experimental values obtained under different conditions. Moreover, typical experimental reporters (Henry's constants and partitioning coefficients) directly probe the excess term calculated by molecular simulations, which is independent of the standard state.

As discussed above, the situation for binding is more complex and requires special care. The first important aspect is that binding free energy calculations must be performed with binding site restraints that implement the binding site indicator function $I(\zeta)$ by forcing the ligand to occupy the binding site even when it is decoupled from the receptor. Doing otherwise would fail to specify the bound macrostate of the complex and return undefined binding free energy estimates. However, in practice, calculations lacking binding site restraints often return reliable binding free energy estimates due to the kinetic trapping of the ligand in the binding site region during the relatively short molecular dynamics sampling typical of alchemical binding free energy simulations. Nevertheless, longer calculations could display persistent drift of the estimate as new regions of conformational space are sampled. A more serious consequence of neglecting binding site restraints is the possibility of the ligand leaving the binding site when it is not coupled to the receptor. Such occurrence indicates a serious violation of the theory. Alchemical binding free energy simulations where the ligand leaves the binding site are not set up correctly and can return grossly inaccurate binding free energy estimates.

Because it is a discontinuous function of the position and orientation of the ligand relative to the receptor, molecular dynamics implementations often approximate the indicator function by means of flat-bottom restraints~\cite{chen2007can, wu2021alchemical} (see below). Harmonic restraints are not as suitable in this respect because they favor a specific microstate at the bottom of the harmonic potential rather than specifying a region of conformational space.

Note that $\Delta G^\circ_{b,{\rm ideal}}$ is independent of the chemical nature of the binding partners, and it is the same for all complexes sharing similar binding site definitions. The latter is the reason that the ideal term cancels out when taking the difference of absolute binding free energies when estimating relative binding free energies (RBFEs). However, while it does not appear explicitly in the definition, $\Delta G_{b,{\rm exc.}}$ depends implicitly on the definition of the binding site through the ensemble average in the uncoupled ensemble, which includes only ``bound'' configurations according to the definition of the indicator function $I(\zeta)$. Varying the binding site definition affects both the ideal and excess binding free energy terms such that their sum (the standard binding free energy) is often only weakly dependent on the binding site definition~\cite{gilson1997statisticalthermodynamic,gallicchio2011recent}. Conversely, great care should be exercised when comparing calculated excess binding free energy values corresponding to different binding site definitions. Similarly, the outcome of a binding free energy calculation conducted without a binding site definition, i.e.\ without binding site restraints, is, in principle, undefined. The latter is also true for relative binding free energies because while the ideal term cancels out, the difference in excess binding free energies depends on the binding pose definition.

\paragraph{Several choices of restraints are possible in binding free energy calculations.}
A variety of restraint types are commonly used in alchemical binding free energy calculations. Before going into their characteristics, we must acknowledge an important and often unrecognized distinction between the binding site restraints corresponding to the indicator function $I(\zeta)$ necessary for the implementation of the theory and optional auxiliary restraints and biasing potentials sometimes used to speed up the convergence of free energy calculations. Binding site restraints are an integral part of the molecular complex definition and are typically maintained throughout the alchemical process. In particular, they must be present unchanged at the unbound and bound states. In contrast, unless thermodynamic reweighting is applied in post-processing, auxiliary restraints cannot be present at the endstates to avoid affecting their free energies. Hence, typically the free energy cost of turning them on is computed at the uncoupled state, and the free energy gain of releasing them is calculated near the bound state. In the alchemical binding free energy literature, often harmonic restraint potentials imposed in an initial stage and removed in a final stage are described as implementing binding sites restraints. However, because they utilize harmonic potentials that cannot correctly describe the binding pose macrostate (see above) and because they are turned on and then off, these are instead auxiliary restraints introduced to enhance convergence without affecting, in principle, binding free energy estimates.

Binding site restraints implement the $I(\zeta)$ indicator function. They, therefore, can depend at most on the six rigid-body external coordinates of the ligand governing the position and orientation of the ligand relative to the receptor. Any other choice would affect the intramolecular conformational distribution of the ligand or the receptor and would  bias the binding free energy estimate. The external degrees of freedom introduced by Boresch et al.~\cite{boresch2003absolute,leitgeb2005alchemical} are a popular choice. The Boresch coordinates, as well as other similar approaches, involve selecting anchor atoms of the receptor and the ligand. In principle, the choice of anchor atoms is arbitrary, provided adequate simulation time is available, and as long as it leads to similar binding pose definitions. However, practical considerations may be important. For example, some choices of Boresch-style anchor atoms can be numerically unstable. Additionally, anchor atoms should likely be placed in a part of the molecule that defines the binding orientation well rather than in a floppy, solvent-exposed tail. Several automated methods have been proposed to select anchor points for Boresch restraints, which typically involve analysing a simulation of the fully-interacting receptor-ligand complex~\cite{alibay2022evaluating, alibay2021mdrestraintsgenerator, baumann2023broadening, chen2023enhancing, hedges2023suite, wu2025optimizing}.

As discussed above, binding site restraints are commonly implemented using quadratic flat-bottom restraints, which impose a restoring harmonic  force similar to harmonic potentials but only if the ligand leaves a specific region~\cite{chen2007can}. Often, binding site restraints do not limit the rotational degrees of freedom of the ligand, and the corresponding standard state term is zero; see Eq.~(\ref{eq:DG0-def-Vsite}). However, it might be challenging to fully explore the bound macrostate at all alchemical states when it is only limited by receptor-ligand positional restraints. As further discussed below, auxiliary restraints are particularly helpful in the latter circumstance to limit the extent of conformational sampling necessary to achieve rapid convergence of binding free energy estimates.

Most current binding free energy workflows do not use binding site restraints present at all stages of the simulation and implicitly assume that $V_{\rm site}$ in Eq.~(\ref{eq:Vsite-def}) is the volume of the simulation box $V$. In the absence of binding site restraints, only the protein-ligand interactions prevent the ligand from slipping out from the binding site and wandering throughout the simulation box within the simulation time frame. However, observations of ligands leaving the binding site become more likely as simulation times increase, especially when analyzing weak binders and non-binders (see below). Moreover, there is no obstacle for the ligand to diffuse away from the binding site at the decoupled state when protein-ligand interactions are turned off.

In practice, because they often implement auxiliary restraints, binding free energy prediction workflows that omit binding site restraints rarely yield grossly erroneous predictions. Auxiliary restraints are introduced gradually as protein-ligand interactions are turned off or are introduced at the bound state when the ligand is naturally trapped in the binding site. Either way, the free energy of imposing them can be tracked while preventing the ligand from leaving the binding site at the decoupled state. Provided that the interacting ligand samples only bound configurations, i.e., it behaves as if binding site restraints were present, the current practice of not using them is likely to incur only minor or insignificant errors.

The Boresch harmonic restraints~\cite{boresch2003absolute,baumann2023broadening} are a popular choice as auxiliary restraints because they are based on interatomic distances, bond angles, and dihedral angles degrees of freedom already implemented in molecular mechanics codes that do not interfere with intramolecular conformational distributions. Moreover, the free energy of releasing Boresch restraints within the simulation box volume $V$ is easily computed analytically~\cite{boresch2003absolute, boresch2024analytical}.

By reducing the conformational space to be sampled, auxiliary restraints are often crucial for obtaining converged binding free energy estimates, even when binding site restraints are employed.
Auxiliary restraints take a variety of forms, such as distance restraints between the ligand and the protein~\cite{mobley2006use,clark2023comparison} as well as flat-bottom restraints. The choice of degrees of freedom and restraining potentials for auxiliary restraints is not as restrictive as for those for binding site restraints, provided that the free energy terms of restraining collective variables that couple protein and ligand degrees of freedom are adequately accounted for to avoid serious systematic errors~\cite{clark2023comparison}.
For example, the overall ligand RMSD has been used~\cite{woo2005calculation} for this purpose. Auxiliary restraints are designed to reduce the amount of conformational sampling at intermediate $\vec{\lambda}$ values. However, more computational effort is required to compute the free energy when the restraints are turned on and off. Additionally, such restraints can limit the exploration of alternative binding modes. This restriction may be undesirable when using Hamiltonian $\lambda$ exchange or expanded ensemble techniques where allowing the ligand to exchange binding modes when it is non-interacting could provide sampling benefits~\cite{wang2013identifying}.
More specifically, flat-bottom restraints allow a ligand to explore multiple binding sites, and harmonic restraints allow the exploration of multiple binding modes within a site, while Boresch restraints only allow a single binding mode within a single site.
See additional discussion of the possibility of multiple binding modes in Sec.~\ref{sec:multiple_binding_modes} below.

Recently, degrees of freedom for binding site and auxiliary potentials based on larger sets of atomic positions of the receptor and ligand have been put forward.
Fu et al. suggested restraining the six relative rigid-body degrees of freedom derived by finding the rotation of the ligand, which minimises the RMSD to a reference structure of the receptor-ligand complex (after accounting for rotation and translation of the receptor)~\cite{fu2017new}. Salari et al. proposed restraining the ``distance-from-bound-configuration'' (DBC) variable, which is the RMSD of ligand coordinates in the frame of reference of the binding site~\cite{salari2018streamlined, ebrahimi2022symmetry}. DBC may allow the binding pose to be closely preserved as the ligand intermolecular interactions are removed, but has the disadvantage of coupling the internal and relative external degrees of freedom of the receptor and ligand, making them unsuitable for describing the binding site macrostate and the corresponding standard state terms. However, these restraints are practical as auxiliary restraints when additional steps are used to introduce and release them. Moreover, because they are based on flat-bottom potentials encompassing the desired binding poses, the free energy of releasing them in the bound state can be safely neglected in most cases. The restraint schemes discussed here are implemented in open-source workflows~\cite{fu2021bfee2, fu2022accurate, santiago-mcrae2023computing, clark2023comparison, hedges2023suite}.

\subsection{Absolute and relative calculations deal with some of the same issues}

\paragraph{Handling weak binders}\label{sec:weak-binders}
According to the non-covalent statistical mechanics theory of bimolecular binding (Sec.~\ref{sec:theory}), only the conformations in which the receptor and ligand form a bound complex are sampled. Determining an appropriate definition of the bound macrostate can be challenging for weakly bound ligands. For tightly bound ligands, virtually all reasonable definitions of the bound state will lead to equivalent free energies since the partition function will be dominated by a relatively limited ensemble of low-energy poses. However, this simplification breaks down for weak binders.  In fact, there might not be a perfect correspondence between the structural definition of the complex on which current alchemical models are based~\cite{gilson1997statisticalthermodynamic} (Sec.~\ref{sec:theory}) and experimental reporters~\cite{mihailescu2004theory}. For example, isothermal titration calorimetry (ITC) or surface plasmon resonance (SPR) measurements effectively define a binding state that encompasses all ligand conformations complexed with the protein, regardless of where they bind the protein. In contrast, fluorescence polarization competition assays measure binding to only a single location, where the ligand of interest displaces a competing binder. Therefore, care must be taken to ensure that a reasonable definition of the binding site is used~\cite{wang2013identifying}.

>>>>

\newcommand{\versionnumber}{2.0} % you should update the minor version number in preprints and major version number of submissions.
\newcommand{\githubrepository}{\url{https://github.com/alchemistry/alchemical-best-practices}} %this should be the main github repository for this article
\newcommand{\expect}[1]{\left\langle{#1}\right\rangle}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]}
\author[1*]{Antonia S. J. S. Mey}
\author[2]{Bryce K. Allen}
\author[3]{Hannah E. Bruce Macdonald}
\author[3*]{John D. Chodera}
\author[9]{Vytautas Gapsys}
\author[9]{David F. Hahn}
\author[1,10]{Maximilian Kuhn}
\author[1]{Julien Michel}
\author[4*]{David L. Mobley}
\author[5]{Levi N. Naden}
\author[6]{Samarjeet Prasad}
\author[2,7]{Andrea Rizzi}
\author[1]{Jenke Scheen}
\author[11]{David R. Slochower}
\author[8*]{Michael R. Shirts}
\author[9]{Gary Tresadern}
\author[2]{Huafeng Xu}

%
\affil[1]{EaStCHEM School of Chemistry, David Brewster Road, Joseph Black Building, The King's Buildings, Edinburgh, EH9 3FJ, UK}
\affil[2]{Silicon Therapeutics, Boston, MA, USA}
\affil[3]{Computational and Systems Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York NY, USA}
\affil[4]{Departments of Pharmaceutical Sciences and Chemistry, University of California, Irvine, Irvine, USA}
\affil[5]{Molecular Sciences Software Institute, Blacksburg VA, USA}
\affil[6]{National Institutes of Health, Bethesda, MD, USA}
\affil[7]{Tri-Institutional Training Program in Computational Biology and Medicine, New York, NY, USA}
\affil[8]{University of Colorado Boulder, Boulder, CO, USA}
\affil[9]{Computational Chemistry, Janssen Research \& Development, Turnhoutseweg 30, Beerse B-2340, Belgium}
\affil[10]{Cresset, Cambridgeshire, UK}
\affil[11]{Vertex Pharmaceuticals, San Diego, CA, USA}

@@

\paragraph{Conserved binding site waters can play an important role in binding free energies}
Binding site water molecules may form water mediated protein-ligand interactions which can pose challenges whenever exchange with bulk water is slow compared to simulation timescales. This can happen in buried binding sites~\cite{laage2017water}. Overlaying multiple protein X-ray structures can identify conserved water molecules or those that are associated only with certain ligands in a series and are therefore important to include in calculations. In cases where water molecules are known to play an important role in ligand binding, for example if the presence of electron density suggests they are tightly bound (indicated by a high EDIA score~\cite{nittinger2015evidence,meyder2017estimating}), software implementations that use water sampling facilitated by Grand Canonical Monte Carlo methods may be useful~\cite{michel2010prediction}. Other tools such as WaterMap~\cite{abel2008role,young2007motifs} or open source equivalents (SSTMap~\cite{haider2018solvation}, GIST~\cite{ramsey2016solvation}, and others) can be used to define the hydration states for systems with no experimental evidence of water sites~\cite{wang2011ligand}. Well-known protein systems with water mediated ligand interactions are for example: HSP90 which formed part of the D3R grand challenge 2015~\cite{mey2016blinded}, A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, ~\cite{deflorian2020accurate}, and others~\cite{michel2009energetics}.

In the context of relative binding free energy calculations, determining the correct hydration pattern for the two endpoints may not be sufficient for an accurate and reproducible calculation. If the locations of the bound water molecules differs at the two endpoints - e.g. if a perturbation to the ligand displaces a bound water molecule - then the network needs to be able to adapt during the perturbation. Otherwise, it is likely that the estimated relative binding free energies will be dependent on the direction in which the perturbation is performed~\cite{ross2020enhancing}.

In absolute binding free energy calculations, a ligand is fully decoupled from a binding site cavity which may otherwise be hydrated in the \textit{apo} state. In this case, the sampling of water molecules must be sufficient to hydrate the binding pocket as the ligand is decoupled (or vice versa for the reverse). When the binding site is exposed and water can move freely between the site and bulk solvent, then sampling is often not an issue. However, in the case of more buried sites or sites which form complex water networks, enhanced water sampling may be required to ensure an accurate binding affinity.

@@

Absolute free energy calculations involve completely removing the interactions between the ligand or solute and its environment, taking it to a non-interacting state that may or may not retain intramolecular non-bonded interactions.
This non-interacting state can then be shifted between environments---from the protein to water, or from one solution to another---without changing its free energy other than that due to the changing volume of the simulations, and then interactions can be restored in the new environment.

Absolute free energies are by definition reported with respect to a specific reference or standard state, which effectively determines the arbitrary point at which the free energy is 0.
The role of the standard state is particularly evident from the expression of the binding free energy between a receptor $R$ and ligand $L$
\begin{equation} \label{eq:DGfromKAB}
    \Delta G = -k_BT ~ \ln \left( c^{\circ} K_b \right)  = -k_BT ~ \ln\left( c^{\circ} \frac{[RL]}{[L][R]} \right) \, .
\end{equation}
Here, the reference state concentration $c^{\circ}$ converts the binding constant $K_b$ into a dimensionless quantity expressed in reference concentration units.
It should be noted that ignoring the term $c^{\circ}$ is equivalent to assuming a reference concentration of 1~D$^{-1}$, where D are the units used to express $K_b$, and would thus cause the value of $\Delta G$ to vary with the choice of the units.
It is convenient to define a standard state at a constant pressure of 1~atm and where each chemical species (i.e., R, L, and RL) in the reaction solvent has a concentration of $c^{\circ}$~=~1~M~=~1~molecule/1660~\r{A}$^3$ but do not interact with other molecules of $R$, $L$, or $RL$.

\paragraph{Handling the standard state in absolute free energy calculations.}
For solvation free energy calculations, handling the standard state is typically straightforward, and treating it correctly simply means ensuring that the non-interacting solute still occupies essentially the same volume as the solute in the interacting system.
So typically in such cases no special care is required to ensure the correct standard state, as long as the \emph{experimental} data being analyzed uses the same standard state.
If this is not the case, a simple entropic correction to the free energy of $k_B T\ln (V_f/V_i) = k_B T \ln (C_i/C_f)$ to the experimental data is needed.

For binding, however, the situation is more complex and requires special care.
Because the simulations are typically performed using restraints and at concentrations that are different from 1 M, the expression of the free energy requires the following correction~\cite{gilson1997statisticalthermodynamic} (see an example of such a thermodynamic cycle in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle})
\begin{equation}\label{eq:restraint-correction}
    \Delta G^{\circ}_{\mathrm{restr}} = -k_BT ~ \ln \left( c^{\circ}{V_L} \right) -k_BT ~ \ln \left( \frac{\xi_L}{8 \pi^2} \right),
\end{equation}
where $V_L$ and $\xi_L$ are respectively the volume of the translational and rotational degrees of freedom of the non-interacting ligand in the simulation box.
When no restraints are used, the non-interacting ligand is free to translate and rotate in the simulation box (i.e., $V_L = V_{\mathrm{box}}$ and $\xi_L = 8\pi^2$), and the rotational term is zero.
A sufficiently thorough exploration of the simulation box by the non-interacting ligand is, however, required for the formula to be valid.
This is typically hard to achieve as the exploration process is governed by diffusion, and weak transient nonspecific binding will occur at other sites on the protein. 
The addition of a restraint limits the volume available to the non-interacting ligand, thus speeding the convergence of the sampling.
In addition, when enhanced sampling methods such as Hamiltonian $\lambda$ exchange are used (see Sec.~\ref{sec:sampling_schemes}), the use of a restraint is typically necessary as it keeps the ligand in the binding site in the interacting state (see also Sec.~\ref{sec:weak-binders}) and generally reduce the round-trip time of replicas.
When restraints are employed, the values of $V_L$ and $\xi_L$ are restraint-dependent, but for commonly employed restraints, these can be usually easily computed analytically or numerically by solving the relevant integral.

\paragraph{Several choices of restraints are possible.}\label{ABFErest}
In practice, a variety of types of restraints are common, from simple harmonic distance restraints between the ligand and the protein~\cite{mobley2006use}, to flat-bottom restraints which work similarly but only exert a force if the ligand leaves a specific region~\cite{chen2007can}.
Because these restraints do not limit the rotational degrees of freedom of the ligand, the rotational term entering the correction in Eq.~\ref{eq:restraint-correction} is zero.

Alternatively, a set of restraints proposed by Boresch have also commonly been employed, where all six rigid-body degrees of freedom governing the orientation of the ligand relative to the receptor are restrained~\cite{boresch2003absolute, leitgeb2005alchemical}.
Further restraints, such as on the overall ligand RMSD have also been used~\cite{woo2005calculation}.

In principle, all of these forms will yield correct binding free energies in the limit of adequate sampling if their effects and connection to the standard state are correctly handled, but they have different strengths and weaknesses.
For example, with more involved restraints, sampling at intermediate $\vec{\lambda}$ values will usually not need to be as extensive but more computational effort must go to computing the free energy to turn on the restraints.
Additionally, such restraints would typically keep the ligand from exploring alternative binding modes. This  restriction may be undesirable when using Hamiltonian $\lambda$ exchange or expanded ensemble techniques where allowing the ligand to exchange binding modes when it is non-interacting could provide sampling benefits~\cite{wang2013identifying}.
More specifically, flat-bottom restraints might allow a ligand to explore multiple binding sites, and harmonic restraints allow exploration of multiple binding modes within a site, while Boresch restraints only allow a single binding mode within a single site.
See additional discussion of the possibility of multiple binding modes in Sec.~\ref{sec:multiple_binding_modes} below.

Many choices of restraints involve selecting reference atoms.
Again, in principle this choice is unimportant given adequate simulation time but practical considerations may be important.
The choice is likely especially important with Boresch-style restraints, where some relative placements of reference atoms are likely to be numerically unstable; additionally, ligand reference atoms should likely be in a part of the molecule which defines the binding orientation well, rather than in a floppy solvent-exposed tail, for example. Several automated methods have been proposed to select anchor points for Boresch restraints, which typically involve analysing a simulation of the fully-interacting receptor-ligand complex~\cite{alibay2022evaluating, alibay2021mdrestraintsgenerator, baumann2023broadening, chen2023enhancing, hedges2023suite}.

More recently, restrictive restraint schemes have been proposed that avoid relying on a few critical reference atoms. Instead, these aim to simplify restraint selection by using larger sets of atomic positions in the receptor and ligand. Fu et al. suggested restraining the six relative rigid-body degrees of freedom derived by finding the rotation of the ligand which minimises RMSD to a reference structure of the receptor-ligand complex (after accounting for rotation and translation of the receptor)~\cite{fu2017new}. Salari et al. proposed restraining the ``distance-from-bound-configuration'' (DBC) variable, which is the RMSD of ligand coordinates in the frame of reference of the binding site~\cite{salari2018streamlined, ebrahimi2022symmetry}. This may allow the binding pose to be closely preserved as the ligand intermolecular interactions are removed, but has the disadvantage of coupling the internal and relative external degrees of freedom of the receptor and ligand, preventing easy computation of the standard state correction. This necessitates an additional step where the DBC restraints are released to a single harmonic restraint. Both restraint schemes are implemented in open-source workflows~\cite{fu2021bfee2, fu2022accurate, santiago-mcrae2023computing}.

\subsection{Absolute and relative calculations deal with some of the same issues}
\subsubsection{Handling weak binders and high dissociation rates}\label{sec:weak-binders}
In binding free energy calculations, only the conformations in which the receptor and ligand form a bound complex should be sampled from the bound states (Sec.~\ref{sec:theory}). Determining what the bound states actually are can be challenging for weakly bound ligands. For tightly bound ligands, virtually all reasonable definitions of the bound state will lead to be equivalent free energies, since the partition function will be dominated by a relatively small number of low-energy poses.

For weak binders, this simplification breaks down.  In fact, the correct bound state may depend on the type of experiment performed.  For example, isothermal titration calorimetry (ITC) or surface plasmon resonance (SPR) measurements effectively define a binding state that includes all ligand comformations that are complexed with the protein, regardless of where on the protein they bind. In contrast, fluorescence polarization competition assays measure binding to only a single location, where the ligand of interest displaces a competing binder. Therefore, care must be taken to ensure that a reasonable definition of the binding site is used~\cite{wang2013identifying}.

In absolute calculations, this need to explicitly define a binding site applies to the fully interacting state in the complex leg of the thermodynamic cycle (top-right state in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}), while in relative calculations the binding site must be defined at both end states of the complex leg (top- and bottom-right states in Fig.~\ref{fig:fig_binding_thermodynamic_cycle}).
In principle, this requires defining which conformations are considered to be bound before running the calculation, but it is common practice to start the simulation with the ligand already placed in the binding site and rely on kinetic trapping to maintain the bound complex.

However, this strategy of using kinetic trapping to maintain the bound complex can fail when the dissociation rate of the ligand has the same or smaller order of magnitude than the length of the simulation.
This is typical of weak binders such as fragments binding shallow pockets with $\mu$M-mM affinities~\cite{georgiou2017pushing,pan2017quantitative}.
In the case of weak binders, using a flat-bottom or harmonic restraint between receptor and ligand in the bound state(s) can prevent dissociations~\cite{georgiou2017pushing,rizzi2019sampl6}.
We stress that this type of restraint is normally avoided as it generally introduces bias in the free energy estimate, which is why the restraint is usually activated only in the intermediate states in absolute calculations.
The bias can be corrected through reweighting schemes~\cite{rizzi2019sampl6}, but this post-processing step can be avoided if a flat-bottom restraint is used and the ligand is never sampled while hitting the potential wall during the simulation in the bound state as the numerical correction will be exactly zero.
It is important to note that the spring constant and/or radius parameters of the restraint effectively determine which conformations are considered to be bound.
As a consequence, these parameters must be tuned to the system so that only the binding site is accessible to the ligand.
Again, this step is particularly important for weak binders as their free energy of binding is known to be more sensitive to the definition of the binding site~\cite{gilson1997statisticalthermodynamic}.

In absolute calculations, this restraint can substitute or be added to the restraint used to handle the standard state correction (Sec.~\ref{sec:standardstate-restraints}).
In the latter case, however, care must be taken when computing the standard state correction.
When multiple restraining potentials are active in the non-interacting state, the correction can generally be computed only through numerical integration.
Alternatively, one can adopt a protocol that removes the bound-state restraint in the non-interacting state.
Finally, even for tight binders, dissociation events can be enhanced by methods such as Hamiltonian replica exchange~\cite{sugita2000multidimensional,chodera2011replica,wang2013identifying} and expanded ensemble~\cite{lyubartsev1992new,li2007simulated}, especially in absolute free energy calculations using harmonic or flat-bottom restraints.
In the latter case, dissociations can be averted simply by increasing the spring constant and/or reducing the radius of the restraint potential to prevent the exploration of ligand conformations outside the binding site in the decoupled state (bottom-right state in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}) that could be propagated to the bound state.
